Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-05-15 papers/ArticleZ80/source/paper.tex
Adicionei as estatisticas de verificacao e fiz algumas correcoes

%[NEW PARAGRAPH-TO-REVIEW]
%Similarly, several other general functions and lemmas were created for all other data types.
%Some times, we need create arithmetic and logic functions that are more specific for a
%determined computer architecture. These functions are developed in the module ALU 
%(arithmetic  logic unit). The following shows some examples of specific functions used in
%microcontroller model of Z80. For example, the function $\textit{instruction_next}$
%receives a $\textit{BYTE}$ and returns a $\textit{BYTE}$ that contains a sequence of bit:
%a signal bit,

Nearly 50\% of all these proof obligations shown in the
Table~\ref{tab:types} are verified automatically or with few proof
commands\footnote{The proof commands are steps that help the prover to find the proof.} .
However many others proof obligations are difficult to verify, as they involve functions
that contain arithmetic expressions that are difficult to manipulate with the interactive
prover of AtelierB. However, all theses 531 proof obligations were verified completely.

@@

This section shows the elements that make up the different types of instructions.
The main elements are specified in the microcontroller module Z80 and parts of it
are presented below.  Basically, the state of microcontroller is formed by data 
from program counter, ports, registers, etc. The transitions between the states 
are actioned by execution of instruction or an external action. Groups of registers
are represented by the variables of the specification states that appear in clause
\textit{VARIABLES}. The declaration of valid states of variables is
in the \textit{INVARIANT} clause and the initial state is defined in the
\textit{INITIALISATION} clause. The assembly instructions are defined by the
clause \textit{OPERATIONS} and, in general, each instruction has three actions: update 
the program counter, update the flag register and its main effect.
%Several functions were created to Z80.
General functions that can be used in other microcontroller models are defined in
the modules of data types. Specific functions of the microcontroller are defined in the ALU (arithmetic logic unit) module.

@@

The interruptions and the \textit{reset} action can change the state of
program counter. Theses actions are modeled by B operations and its main
effects are presented here\footnote{Some definitions of constants: $\textit{sp\_minus\_two}$ holds the value of stack pointer minus 2,
%=$\textit{dec\_BV16(dec\_BV16(sp))}$
 $\textit{sp\_minus\_one}$ is the value of stack pointer minus 1,
%=$\textit{dec\_BV16(sp)}$
$\textit{pc\_high}$ holds the most significant 8 bits and
$\textit{pc\_low}$ holds the least significant 8 bits.}.

@@

\subsection{Verification process}% ESTOU TRABALHANDO AQUI

% parágrafo inicial
This project has many proof obligations and its construction of then model was a difficult task, as several
% iterations were needed to provide the good library definitions
% as well as to optimize the definition of the functionality of
% the microprocessor instructions by grouping common data
% manipulation into auxiliary functions.

% [O que as provas garantem? citar que dificeis, que é necessário buscar técnicas eficientes
% para realizar as provas ]

The proof obligations allow to verify the data types, important system
properties and if the expressions are well-defined (WD)\footnote{An
  expression is called ``well-defined'' (or unambiguous) if its
  definition assigns it a unique interpretation or value.}. The
properties provide additional guarantees, because they can set many
safety rules. However, the model can be very difficult to prove.

% Conclusão: toque final
Several iterations were needed to provide the good library definitions as well as
to fine-tune the model of the microcontroller instructions by factoring common
functionalities into auxiliary definitions.

%OK -> como as regras foram utilizadas ?  Respondido na seção das regras -

%-> Quanto tempo foi gasto nessa verificação ? a minha dissertação diz que aproximadamente dois meses. Utilizando as regras este tempo caiu p/ 5 dias (linkar com este parágrafo)

%-> Aproximadamente quanto tempo de processamento elas reduziram ?
% Porque o modelo não foi verificado completamente ? Isto é viavel ?
 %There are many benefits when a formal model is
%developed, but to verify the model is not an easy task, the first version of Z80 model have already verified %completely,but the new Z80 model with support to animation in ProB has not been verified completely,
%because the changes in Z80 model are recent and the verification time is long.

TODO:ENGLISH REVIEW

%Decomposição do módulo Z80
The Z80 module was decomposed in 8 modules of instructions, it was very helpful to decrease the time verification, because the AtelierB can start the simultaneously 8 instances of provers reducing the time verification by up to 8 times, when it uses a processor with 8 cores.

 %-> O que significa cada coluna e linha da tabela ?  	
The follow table~\ref{tab:completestatistcs} contains the results of verification process, it shows only the number of no obvious proof obligations and the titles of columns use acronyms: {\small \textbf{PO}} - number of proof obligations; {\small \textbf{UPO}} - number of unproved proof obligations; {\small \textbf{WD PO}} - number of proof obligations  related to well definition lemmas; {\small \textbf{UWD PO}} - number of unproved proof obligations related to well definition lemmas.

@@

% Resumo de como aconteceu o processo de verificação, (melhor está após as estátisticas) ( )Atualizar dados
However, few proof commands need to be used to prove most proof obligations. As there are many similar assembly instructions,
some human-directed proofs, when replayed, could discharge other proof
obligations. A good example is a set of 17 proof commands that quickly aided the
verification of 99\% (2295) of WD proofs. We also set up a proving environment
consisting of networked computers to take advantage of the distribution
facilities now provided in the B development environment. Finally, all of the
2926 proof obligations were proved using the tool support of the development
environment.

@@

@@

>>>>

%TODO:[NEW PARAGRAPH-TO-REVIEW]
Similarly, several other general functions and lemmas were created for all other data types.
Some times, we need  create arithmetic and logic functions that are more specific for a
determined microprocessor, like Z80 that is presented in the section \ref{sec:z80}.

%TODO:[NEW PARAGRAPH-TO-REVIEW]
The definitions, functions and properties from hardware library and types library are difficult to verify,
on the other hand  these libraries can be reused in several others  microprocessors without to verify again.
Nearly 50\% of all these proof obligations shown in the Table~\ref{tab:types} are verified automatically 
or with few user pass\footnote{The user pass are steps that aid the prover to find the proof.} .
But many others proof obligations are difficult to verify, because they involve functions
that contain arithmetic expressions that are complex to manipulate with the interactive
prover of AtelierB.  However, all theses 531 proof obligations were verified completely.

@@

This section shows the elements that make up the different types of
instructions. The main elements are specified in the microcontroller module Z80
and parts of it are presented below.  Basically, the state of microcontroller is
formed by data from program counter, ports, registers, etc. The transitions
between the states are actioned by execution of instruction or an external
action. Groups of registers are represented by the variables of the
specification states that appear in clause \textit{VARIABLES}. The declaration
of valid states of variables is in the \textit{INVARIANT} clause and the initial
state is defined in the \textit{INITIALISATION} clause. The assembly
instructions are defined by the clause \textit{OPERATIONS} and, in general, each
instruction has three actions: update the program counter, update the flag
register and its main effect.
% Several functions were created to Z80.
General functions that can be used in other microcontroller models are defined
in the modules of data types. Specific functions of the microcontroller are
defined in the ALU (arithmetic logic unit) module.

@@

The interruptions and the \textit{reset} action can change the state of program
counter. Theses actions are modeled by B operations and its main effects are
presented here\footnote{Some definitions of constants: $\textit{sp\_minus\_two}$
holds the value of stack pointer minus 2,
% =$\textit{dec\_BV16(dec\_BV16(sp))}$
 $\textit{sp\_minus\_one}$ is the value of stack pointer minus 1,
% =$\textit{dec\_BV16(sp)}$
$\textit{pc\_high}$ holds the most significant 8 bits and $\textit{pc\_low}$
holds the least significant 8 bits.}.

@@

\subsection{Verification process}% ESTOU TRABALHANDO AQUI
% OK -> como as regras foram utilizadas ?  Respondido na seção das regras
% TODO:ENGLISH REVIEW parágrafo inicial: VISÃO GERAL
This project has many proof obligations and its verification can be a hard task,
hence this section explains about how happen the verification process of proof
obligations and its importance. % [O que as provas garantem? citar que énecessário buscar técnicas eficientes para realizar as provas ]
The proof obligations allow to verify the data types, important system properties and if
the expressions are well-defined (WD)\footnote{An expression is called ``well-defined''
(or unambiguous) if its definition assigns it a unique interpretation or value.}. The
properties provide additional guarantees, because they can set many safety
rules. However, the model can be very difficult to prove.

 % -> O que significa cada coluna e linha da tabela ?
The Table~\ref{tab:completestatistcs} contains some statistics of verification
process, it shows only the number of no obvious proof
obligations\footnote{Obvious proof obligations are simple formulas automatically eliminated by
the tool.} and the titles of columns use acronyms: {\small \textbf{PO}} - number
of proof obligations; {\small \textbf{UPO}} - number of unproved proof
obligations; {\small \textbf{WD PO}} - number of proof obligations  related to
well definition lemmas; {\small \textbf{UWD PO}} - number of unproved proof
obligations related to well definition lemmas.

@@

%->Explicaçao de mais detalhes da tabela.

% Porque o modelo não foi verificado completamente ?
There are many benefits when a formal model is developed, but to verify the
model is not an easy task, the first version of Z80 model have already verified
completely, however this new Z80 model with support to animation in ProB has not
been verified completely, because the changes in Z80 model are recent and the
verification time is long.
% ->Quanto tempo foi gasto nessa verificação ?  Aproximadamente quanto tempo de
% processamento elas reduziram ?
Nevertheless this project has 3631 proof obligations and the first version of
Z80 model spent approximately two months of work to finish the verification
process. The same  developed verification strategies were adjusted and reused in
new Z80 model, so approximately  7 days of work was enough to verify 3062  proof
obligations (84\%) and remaining only 569 proof obligations (16\%) not yet
verified.

Two development strategies were important to improve the verification of 3062
proof obligations: the intensive use of parallelism and user pass. %Decomposição
do módulo Z80 The Z80 module was decomposed in 8 modules of instructions, it was
very helpful to decrease the time for verification, because AtelierB can start 
simultaneously \textit{N} instances of provers, one per module, reducing the
time verification by up to \textit{N}  times when a processor with \textit{N} 
cores is used  or the process is submitted to \textit{N} networked computers. So
we also set up a proving environment consisting of networked computers to take
advantage of the distribution facilities provided in the B development
environment.
% Resumo de como aconteceu o processo de verificação,
A second interesting strategy  was to reuse the user pass. Usually, an used
user pass to aid the verification of one proof obligation can solve several
similar proof obligations. Thus few generic user pass are needed to prove many
proof obligations. As there are many similar assembly instructions, some
human-directed proofs, when replayed, could discharge other proof obligations. A
good example is a set of 17 user pass that quickly aided the verification of
approximately 2533 of WD proofs.

@@

%\section{Case study}
%\include{Petroleumproductiontestsystem}

@@

% Conclusão: toque final
%Several iterations were needed to provide the good library definitions as well as
%to fine-tune the model of the microcontroller instructions by factoring common
%functionalities into auxiliary definitions.
